Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-03-02 doc/paper/paper.tex
Clean up the Preliminaries

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Vone}{\mathsf{V}_{1}}
\newcommand{\Vtwo}{\mathsf{V}_{2}}
\newcommand{\Vthree}{\mathsf{V}_{3}}

@@

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. Despite
  being around for more than a decade, practical FOMC algorithms are still
  unable to compute functions as simple as a factorial. We argue that the
  capabilities of FOMC algorithms to date are limited by their inability to
  express arbitrary recursive computations. To enable arbitrary recursion, we
  relax the restrictions that typically accompany domain recursion and
  generalise circuits used to express a solution to an FOMC problem to graphs
  that may contain cycles. To this end, we enhance the most well-established
  (weighted) FOMC algorithm \textsc{ForcLift} with new compilation rules and an
  algorithm to check whether a recursive call is feasible. These improvements
  allow the algorithm to construct efficient solutions to counting fundamental
  structures such as injections and bijections. We end with a few conjectures on
  what classes of instances could be liftable as a result.
\end{abstract}

% TODO: I found the choice for what goes into a footnote and what doesn't rather
% inconsistent, which made it a bit irritating to read for me. Maybe review if
% the current use of footnotes is really necessary and helpful?

@@

% TODO: Maybe some of the info from the preliminaries is redundant now that I've
% moved the proofs elsewhere?

% TODO: I find the writing mostly clear, but the structure of the Preliminaries
% somewhat confusing. It starts with some notation, then some definitions, then
% suddenly a paragraph titled "Notation", and so on. Maybe see if you can
% rearrange some things and/or add some (more) helpful paragraph titles?

@@

@@

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}
  is a triple $c = (L, E, \delta_c)$, where $L$ is a set of literals, $E$ is a
  set of constraints, and $\delta_c$ is the domain map of $c$. Let $\Vars$ be
  the function that maps clauses and sets of either literals or constraints to
  the set of variables in them. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(E)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in E$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

% TODO: The first paragraph already introduced some notation.
\paragraph{Notation.}
For any set $S$, let $2^{S}$ denote its power set and
$S^{\ast} \coloneqq \bigcup_{i=0}^{\infty} S^{i}$ the set of tuples of any
finite length of elements of $S$. For example,
$\{\,a, b\,\}^{\ast} = \{\, (), (a), (b), (a, a), \dots \,\}$. For any
$n \in \mathbb{N}_{0}$, let $[n] \coloneqq \{\, 1, \dots, n \,\}$, e.g.,
$[0] = \emptyset$, and $[2] = \{\, 1, 2 \,\}$. We write $\langle\rangle$ for an
empty list, $\langle x \rangle$ for a list with one element $x$, $|l|$ for the
length of list $l$, $\mdoubleplus$ for list concatenation, and $\pfun$ to denote
partial functions. Let $S$ be a set of constraints or literals, $V$ a set of
variables, and $x$ a variable or a constant. We write $S[x/V]$ to denote $S$
with all occurrences of all variables in $V$ replaced with $x$. Let $\Doms$
(respectively, $\Preds$) be the function that maps any clause or formula to the
set of domains (respectively, predicates) used within, and let $\dom(f)$ denote
the domain of function $f$.

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} is a set of clauses s.t.\ all constraints and atoms `type
  check' with respect to domains. Each formula $\phi$ has a map
  $\pi_{\phi}\colon \Preds(\phi) \to {\Doms(\phi)}^{\ast}$ such that, for each
  predicate $\texttt{p}/n \in \Preds(\phi)$, we have that
  $\pi_{\phi}(\texttt{p}) \in {\Doms(\phi)}^{n}$.
\end{definition}

\begin{remark}
  Constants are also mapped to domains, although we keep this mapping implicit.
\end{remark}

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $\texttt{p}/2$, variables $X$, $Y$, $Z$, and domains
  $\Gamma$ and $\Delta$. Based on $\delta_{c_{1}}$ and $\delta_{c_{2}}$, we can
  infer that $\pi_{\phi} = \{\, \texttt{p} \mapsto (\Gamma, \Delta) \,\}$. All
  variables that occur as the first argument to $\texttt{p}$ are in $\Gamma$,
  and, likewise, all variables that occur as the second argument to $\texttt{p}$
  are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

@@

% TODO: the first sentence comes out of nowhere
Once domains are mapped to finite sets and constants to elements in those sets,
a formula can be viewed as a set of conditions that the predicates (interpreted
as relations) have to satisfy. Hence, FOMC is the problem of counting the number
of combinations of relations that satisfy these conditions.

@@

To formalise the idea of a model, let
$\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size function}
that maps each domain to a non-negative integer.

\begin{definition}\label{def:model}
  Let $\phi$ be a formula and $\sigma$ a domain size function. A \emph{model} of
  $(\phi, \sigma)$ is a map
  $\mathfrak{M}\colon \Preds(\phi) \to 2^{\mathbb{N}_{0}^{\ast}}$ s.t.\ the
  following two conditions are satisfied.
  \begin{enumerate}
    \item Let $\texttt{p}/n \in \Preds(\phi)$ be a predicate and let
          $\pi_{\phi}(\texttt{p}) = {(d_{i})}_{i=1}^{n}$ for some domains
          $d_{i} \in \Doms(\phi)$. Then
          \begin{equation}\label{eq:subset}
            \mathfrak{M}(\texttt{p}) \subseteq \prod_{i=1}^{n}[\sigma(d_{i})].\footnote{For simplicity, \cref{def:model} ignores constants. To include constants, one would replace $[\sigma(d_{i})]$ with a set that contains all constants associated with domain $d_{i}$, extended with enough new elements to make its cardinality $\sigma(d_{i})$.}
          \end{equation}
    \item As a collection of relations, $\mathfrak{M}$ satisfies $\phi$.
    \begin{itemize}
      \item TODO: introduce the substitution for formulas, moving the definition from elsewhere
      \item define inductively as...
      \item (not exhaustive)
      \item $\mathfrak{M} \models \phi \land \psi$ iff $\mathfrak{M} \models \phi$ and $\mathfrak{M} \models \psi$
      \item $\mathfrak{M} \models \phi \lor \psi$ iff $\mathfrak{M} \models \phi$ or $\mathfrak{M} \models \psi$
      \item $\mathfrak{M} \models \neg\phi$ iff $\mathfrak{M} \not\models \phi$
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{. }\phi)$ iff $\mathfrak{M} \models \phi[x/\{\, X \,\}]$ for all $x \in \Delta$
      \item One can then add inequality constraints to the semantics as follows
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{. } X \ne c \Rightarrow \phi)$ iff $\mathfrak{M} \models \phi[x/\{\, X \,\}]$ for all $x \in \Delta \setminus \{\, c \,\}$
      \item $\mathfrak{M} \models (\forall X,Y \in \Delta\text{. } X \ne Y \Rightarrow \phi)$ iff $\mathfrak{M} \models \phi[x/\{\, X \,\}][y/\{\, Y \,\}]$ for all $x,y \in \Delta$ such that $x \ne y$
    \end{itemize}
  \end{enumerate}
\end{definition}

% TODO: make the relations more concrete
\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\sigma(\Gamma) = \sigma(\Delta) = 2$. Then the right-hand side of
  \cref{eq:subset} becomes
  \[
    [\sigma(\Gamma)] \times [\sigma(\Delta)] = [2] \times [2] = \{\,(1, 1), (1, 2), (2, 1), (2, 2)\,\}
  \]
  There are $2^{4} = 16$ possible relations between $\Gamma$ and $\Delta$. Let
  us count how many of them satisfy the conditions imposed on predicate
  $\texttt{p}$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $(\phi, \sigma)$ is 7. Incidentally, it
  counts partial injections. We will continue to use the problem of counting
  partial injections (and the formula from \cref{example:first} specifically) as
  the main running example throughout the paper.
\end{example}

@@

  \SetKwData{functionCall}{call}
  \SetKwData{functionSignature}{signature}
  \SetKwData{functionNames}{F}
  \SetKwData{variableNames}{V}
  \SetKwData{newVariableNames}{V'}
  \SetKwData{expression}{e}
  \SetKwData{functions}{funs}
  \SetKwData{args}{args}

  \SetKwFunction{newFunctionName}{newFunctionName}
  \SetKwFunction{newVariableName}{newVarName}
  \SetKwFunction{visit}{v}
  \SetKwFunction{actuallyVisit}{visit}

  \KwIn{$\mathcal{D}$---the set of domains of the input formula}
  \KwIn{$s$---the source node of the FCG}
  \KwData{$\functionNames = \emptyset$ (function names)}
  \KwOut{a list of function definitions}
  $\variableNames \gets \{\, \Omega \mapsto \newVariableName{} \mid \Omega \in \mathcal{D} \,\}$\;\label{line:varnames}
  \If{$s$ is a direct successor of a $\Reff$ node}{
    $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;\label{line:initialcall}
    \Return{\functions}\;\label{line:initialcalltwo}
  }
  $f \gets \newFunctionName{$s$}$\;
  $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;
  \Return{$\langle f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathcal{D} \rangle) = \expression \rangle \mdoubleplus \functions$}\;
  \Fn{\visit{node $v$, variable names \variableNames}}{
    \If{$v$ is \textbf{not} a direct successor of a $\Reff$ node}{
      \Return{\actuallyVisit{$v$, \variableNames}}\;
    }
    $f \gets \newFunctionName{$v$}$\;
    $\newVariableNames \gets \variableNames$\;
    \For{$\Omega \in \mathfrak{D}(v)$ s.t. $\newVariableNames(\Omega)$ is non-atomic}{\label{line:nonone}
      $\newVariableNames(\Omega) \gets \newVariableName{}$\;\label{line:nontwo}
    }
    $(\expression, \functions) \gets \actuallyVisit{$v$, \newVariableNames}$\;\label{line:actuallyvisit}
    $\functionCall \gets f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    $\functionSignature \gets f(\langle \newVariableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    \Return{$(\functionCall, \langle \functionSignature = \expression \rangle \mdoubleplus \functions)$}\;
  }
  \Fn{\actuallyVisit{node $v$, variable names \variableNames}}{
    \Switch{label of $v$}{
      \lCase{$\GDR(v')$}{\Return{\visit{$v'$, \variableNames}}}
      \uCase{$\CR_{\Omega \mapsto \Omega'}(v')$}{
        \Return{\visit{$v'$, $\variableNames \cup \{\, \Omega' \mapsto (\variableNames(\Omega) - 1) \,\}$}}\;\label{line:cr}
      }
      \uCase{$\Reff_{\rho}(v')$}{
        $\args \gets \langle \variableNames\circ\rho(\Omega) \mid \Omega \in \mathfrak{D}(v') \rangle$\;\label{line:args}
        \Return{$(\functionNames{$v'$}(\args), \langle \rangle)$}\;
      }
      \dots
    }
  }
\end{algorithm}

\begin{itemize}
  \item TODO: is maybe `parameter name' (or something similar) would be more
        appropriate than `variable name'?
  \item We call a variable name is \emph{atomic} if it is free of subtraction
        (e.g., $m$, $n$); otherwise it is \emph{non-atomic} (e.g., $m-1$,
        $n-l$).
  \item We assume a fixed total ordering of all domains names. In particular, in
        our running example we assume that $\Gamma$ goes before $\Delta$.
  \item For each node $v$, let $\mathfrak{D}(v)$ denote the (precomputed) list
        of domains, the sizes of which we would use as parameters if we were to
        define a function that begins at $v$. As a set, it is computed by
        iteratively setting
        $\mathfrak{D}(v) \gets \left(\bigcup_{u} \mathfrak{D}(u) \setminus I_{v}\right) \cup U_{v}$
        until convergence, where:
        \begin{enumerate*}[label=(\roman*)]
          \item the union is over all direct successors of $v$,
          \item $I_{v}$ is the set of domains \emph{introduced} at node $v$, and
          \item $U_{v}$ is the set of domains \emph{used} by $v$.
        \end{enumerate*}
        The set is then sorted according to the fixed ordering of domains.
  \item $\mathsf{F}$ is a partial map from the set of FCG nodes to function
        names (e.g., $f$, $g$).
  \item Functions \newVariableName and \newFunctionName generate
        previously-unused names for variables and functions, respectively. The
        latter also records the new name in $\mathsf{F}$.
  \item Note that a node is a direct successor of a $\Reff$ node if and only if
        it has an in-degree greater than one.
  \item For simplicity, we refer to FCG nodes using their types.
  \item The algorithm consists of two main functions: \visit and \actuallyVisit.
        The former handles new function definitions, and the latter produces an
        algebraic interpretation of each node depending on its type. As there
        are many node types, \cref{alg:interpretation} only includes the ones
        pertinent to the contributions of this paper; see previous
        work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information about the other
        types.
  \item Given some node $v$ as input, both \visit and \actuallyVisit return a
        pair $(\mathsf{e}, \mathsf{funs})$. Here, $\mathsf{e}$ is the algebraic
        expression that should be used to evaluate $v$ or use it in another
        expression, and $\mathsf{funs}$ is a list of auxiliary functions that
        were created while constructing $\mathsf{e}$.
\end{itemize}

@@

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. The interpretation of a $\CR$ or a $\GDR$ node is simply the
interpretation of its only direct successor. Obviously, $\star$ nodes have no
interpretation as incomplete FCGs are not meant to be interpreted. The
interpretation of a $\Reff$ node is a function call. The direct successor of the
$\Reff$ node (say, $v$) then must introduce a function. The parameters of this
function are the sizes of all domains used by nodes reachable from $v$.

\begin{figure}
  \centering
  \begin{tikzcd}
    \ar[d] & \\
    \text{\Cref{alg:interpretation}} \ar[d] \ar[u, swap, shift right=0.7ex, "{\left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}"] & \\
    \visit{$\GDR$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({f(m, n), \left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}\right)"] & \\
    \actuallyVisit{$\GDR$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bigvee$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\CR$, $\Vtwo$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\land$, $\Vthree$} \ar[d] \ar[dr] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bot$, $\Vthree$} \ar[u, swap, shift right=0.7ex, "\left({[l<2], \langle\rangle}\right)"] & \visit/\actuallyVisit{$\Reff$, $\Vthree$} \ar[ul, swap, shift right=0.7ex, "\left({f(m-1, n-l), \langle\rangle}\right)"]
  \end{tikzcd}
  \caption{Function calls and their return values when \cref{alg:interpretation}
    is run on the FCG from \cref{fig:examplefcg}. When \protect\visit calls
    \protect\actuallyVisit and returns its output unchanged, the two function
    calls are shortened as a single call to
    $\protect\visit/\protect\actuallyVisit$. TODO: this last sentence should be
    explained in text.}\label{fig:functioncalls}
\end{figure}

\begin{example}\label{example:interpretation}
  Let us see how \cref{alg:interpretation} handles the FCG from
  \cref{fig:examplefcg}. Here, $\mathcal{D} = \{\, \Gamma, \Delta \,\}$. Let
  $\Vone \coloneqq \{\, \Gamma \mapsto m, \Delta \mapsto n \,\}$,
  $\Vtwo \coloneqq \Vone \cup \{\, \Delta^{\top} \mapsto l, \Delta^{\bot} \mapsto (n-l) \,\}$,
  and $\Vthree \coloneqq \Vtwo \cup \{\, \Gamma' \mapsto (m-1) \,\}$ denote
  versions of $\mathsf{V}$ at various points throughout the algorithm's
  execution. Here, $l$, $m$, and $n$ are all arbitrary variable names generated
  by \newVariableName. \Cref{line:varnames} initialises $\mathsf{V}$ to $\Vone$.

>>>>

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Done}{\mathsf{D}_{1}}
\newcommand{\Dtwo}{\mathsf{D}_{2}}
\newcommand{\Dthree}{\mathsf{D}_{3}}

@@

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. In this
  paper, we argue that the capabilities of FOMC algorithms to date are limited
  by their inability to express many types of recursive computations. To enable
  such computations, we relax the restrictions that typically accompany domain
  recursion and generalise the circuits used to express a solution to an FOMC
  problem to directed graphs that may contain cycles. To this end, we adapt the
  most well-established (weighted) FOMC algorithm \textsc{ForcLift} to work with
  such graphs and introduce new compilation rules that can create cycle-inducing
  edges that encode recursive function calls. These improvements allow the
  algorithm to find efficient solutions to counting problems that were
  previously beyond its reach, including those that cannot be solved efficiently
  by any other exact FOMC algorithm. We end with a few conjectures on what
  classes of instances could be domain-liftable as a result.
\end{abstract}

% KEYWORDS:
% model counting
% first-order logic
% recursion
% knowledge compilation

% TOPICS:
% KR & ML: probabilistic reasoning and learning
% KR & ML: statistical relational learning
% (no) computational aspects of knowledge representation
% (no) knowledge representation languages
% (no) KR & ML: graphical models for knowledge representation and reasoning

@@

@@

We use $\langle\cdot\rangle$ to denote lists, $|\cdot|$ for the length of a
list, and $\mdoubleplus$ for list concatenation. We write $\pfun$ to denote
partial functions and $\dom(\cdot)$ for the domain of a function. Let $\Doms$ be
the function that maps any clause or formula to the set of domains used within.
Similarly, let $\Vars$ be the function that maps clauses and sets of either
literals or constraints to the set of variables in them. Let $S$ be a set of
constraints or literals, $V$ a set of variables, and $x$ a variable or a
constant. We write $S[x/V]$ to denote $S$ with all occurrences of all variables
in $V$ replaced with $x$.

@@

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause} is a triple $c = (L, E, \delta_c)$, where $L$ is a set of
  literals, $E$ is a set of constraints, and $\delta_c$ is the domain map.
  \emph{Domain map} $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function
  that maps all variables in $c$ to their domains such that (s.t.) if
  $(X, Y) \in E$ for some variables $X$ and $Y$, then
  $\delta_c(X) = \delta_c(Y)$. For convenience, we sometimes write $\delta_c$
  for the domain map of $c$ without unpacking $c$ into its three constituents.
\end{definition}

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} is a set of clauses s.t.\ all constraints and atoms `type
  check' with respect to domains.
\end{definition}

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $\texttt{p}/2$, variables $X$, $Y$, $Z$, and domains
  $\Gamma$ and $\Delta$. All variables that occur as the first argument to
  $\texttt{p}$ are in $\Gamma$, and, likewise, all variables that occur as the
  second argument to $\texttt{p}$ are in $\Delta$. Therefore, $\phi$ `type
  checks' as a valid formula.
\end{example}

@@

@@

The model count of a formula $\phi$ also depends on the sizes of the domains in
$\phi$. Let $\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain
  size function} that maps each domain to a non-negative integer.

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\Gamma = \Delta = \{\,1, 2\,\}$, i.e., $\sigma(\Gamma) = \sigma(\Delta) = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $\texttt{p}$. The empty relation does. All four relations of
  cardinality one (e.g., $\{\, (1, 1) \,\}$) do too. Finally, there are two
  relations of cardinality two ($\{\, (1, 1), (2, 2) \,\}$ and
  $\{\, (1, 2), (2, 1) \,\}$) that satisfy the conditions as well. Thus, the
  FOMC of $(\phi, \sigma)$ is 7. Incidentally, the FOMC of $\phi$ counts partial
  injections from $\Gamma$ to $\Delta$. We will continue to use the problem of
  counting partial injections (and the formula from \cref{example:first}
  specifically) as the main running example.
\end{example}

@@

  \SetKwData{functionCall}{call}
  \SetKwData{functionSignature}{signature}
  \SetKwData{functionNames}{F}
  \SetKwData{domainNames}{D}
  \SetKwData{newDomainNames}{D'}
  \SetKwData{expression}{e}
  \SetKwData{functions}{funs}
  \SetKwData{args}{args}

  \SetKwFunction{newFunctionName}{newFunctionName}
  \SetKwFunction{newDomainName}{newDomainName}
  \SetKwFunction{visit}{v}
  \SetKwFunction{actuallyVisit}{visit}

  \KwIn{$\mathcal{D}$---the set of domains of the input formula}
  \KwIn{$s$---the source node of the FCG}
  \KwData{$\functionNames = \emptyset$ (function names)}
  \KwOut{a list of function definitions}
  $\domainNames \gets \{\, \Omega \mapsto \newDomainName{} \mid \Omega \in \mathcal{D} \,\}$\;\label{line:varnames}
  \If{$s$ is a direct successor of a $\Reff$ node}{
    $(\expression, \functions) \gets \visit{$s$, $\domainNames$}$\;\label{line:initialcall}
    \Return{\functions}\;\label{line:initialcalltwo}
  }
  $f \gets \newFunctionName{$s$}$\;
  $(\expression, \functions) \gets \visit{$s$, $\domainNames$}$\;
  \Return{$\langle f(\langle \domainNames{$\Omega$} \mid \Omega \in \mathcal{D} \rangle) = \expression \rangle \mdoubleplus \functions$}\;
  \Fn{\visit{node $v$, domain names \domainNames}}{
    \If{$v$ is \textbf{not} a direct successor of a $\Reff$ node}{
      \Return{\actuallyVisit{$v$, \domainNames}}\;
    }
    $f \gets \newFunctionName{$v$}$\;
    $\newDomainNames \gets \domainNames$\;
    \For{$\Omega \in \mathfrak{D}(v)$ s.t. $\newDomainNames(\Omega)$ is non-atomic}{\label{line:nonone}
      $\newDomainNames(\Omega) \gets \newDomainName{}$\;\label{line:nontwo}
    }
    $(\expression, \functions) \gets \actuallyVisit{$v$, \newDomainNames}$\;\label{line:actuallyvisit}
    $\functionCall \gets f(\langle \domainNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    $\functionSignature \gets f(\langle \newDomainNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    \Return{$(\functionCall, \langle \functionSignature = \expression \rangle \mdoubleplus \functions)$}\;
  }
  \Fn{\actuallyVisit{node $v$, domain names \domainNames}}{
    \Switch{label of $v$}{
      \lCase{$\GDR(v')$}{\Return{\visit{$v'$, \domainNames}}}
      \uCase{$\CR_{\Omega \mapsto \Omega'}(v')$}{
        \Return{\visit{$v'$, $\domainNames \cup \{\, \Omega' \mapsto (\domainNames(\Omega) - 1) \,\}$}}\;\label{line:cr}
      }
      \uCase{$\Reff_{\rho}(v')$}{
        $\args \gets \langle \domainNames\circ\rho(\Omega) \mid \Omega \in \mathfrak{D}(v') \rangle$\;\label{line:args}
        \Return{$(\functionNames{$v'$}(\args), \langle \rangle)$}\;
      }
      \dots
    }
  }
\end{algorithm}

\begin{itemize}
  \item We call a domain name \emph{atomic} if it is free of subtraction (e.g.,
        $m$, $n$); otherwise it is \emph{non-atomic} (e.g., $m-1$, $n-l$).
  \item We assume a fixed total ordering of all domains. In particular, in our
        running example we assume that $\Gamma$ goes before $\Delta$.
  \item For each node $v$, let $\mathfrak{D}(v)$ denote the (precomputed) list
        of domains, the sizes of which we would use as parameters if we were to
        define a function that begins at $v$. As a set, it is computed by
        iteratively setting
        $\mathfrak{D}(v) \gets \left(\bigcup_{u} \mathfrak{D}(u) \setminus I_{v}\right) \cup U_{v}$
        until convergence, where:
        \begin{enumerate*}[label=(\roman*)]
          \item the union is over all direct successors of $v$,
          \item $I_{v}$ is the set of domains \emph{introduced} at node $v$, and
          \item $U_{v}$ is the set of domains \emph{used} by $v$.
        \end{enumerate*}
        The set is then sorted according to the fixed ordering of domains.
  \item $\mathsf{F}$ is a partial map from the set of FCG nodes to function
        names (e.g., $f$, $g$).
  \item Functions \newDomainName and \newFunctionName both generate
        previously-unused names. The latter also records the new name in
        $\mathsf{F}$.
  \item Note that a node is a direct successor of a $\Reff$ node if and only if
        it has an in-degree greater than one.
  \item The algorithm consists of two main functions: \visit and \actuallyVisit.
        The former handles new function definitions, and the latter produces an
        algebraic interpretation of each node depending on its type. As there
        are many node types, \cref{alg:interpretation} only includes the ones
        pertinent to the contributions of this paper; see previous
        work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information about the other
        types.
  \item Given some node $v$ as input, both \visit and \actuallyVisit return a
        pair $(\mathsf{e}, \mathsf{funs})$. Here, $\mathsf{e}$ is the algebraic
        expression that should be used to evaluate $v$ or use it in another
        expression, and $\mathsf{funs}$ is a list of auxiliary functions that
        were created while constructing $\mathsf{e}$.
\end{itemize}

@@

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. The interpretation of a $\CR$ or a $\GDR$ node is simply the
interpretation of its only direct successor. The interpretation of a $\Reff$
node is a function call. The direct successor of the $\Reff$ node (say, $v$)
then must introduce a function. The parameters of this function are the sizes of
all domains used by nodes reachable from $v$.

\begin{figure}
  \centering
  \begin{tikzcd}
    \ar[d] & \\
    \text{\Cref{alg:interpretation}} \ar[d] \ar[u, swap, shift right=0.7ex, "{\left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}"] & \\
    \visit{$\GDR$, $\Done$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({f(m, n), \left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}\right)"] & \\
    \actuallyVisit{$\GDR$, $\Done$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bigvee$, $\Done$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\CR$, $\Dtwo$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\land$, $\Dthree$} \ar[d] \ar[dr] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bot$, $\Dthree$} \ar[u, swap, shift right=0.7ex, "\left({[l<2], \langle\rangle}\right)"] & \visit/\actuallyVisit{$\Reff$, $\Dthree$} \ar[ul, swap, shift right=0.7ex, "\left({f(m-1, n-l), \langle\rangle}\right)"]
  \end{tikzcd}
  \caption{Function calls and their return values when \cref{alg:interpretation}
    is run on the FCG from \cref{fig:examplefcg}.}\label{fig:functioncalls} % TODO: do I need a dot?
\end{figure}

% TODO: examine the order in which everything is introduced
\begin{example}\label{example:interpretation}
  Let us see how \cref{alg:interpretation} handles the FCG from
  \cref{fig:examplefcg}. In this case, $\mathcal{D} = \{\, \Gamma, \Delta \,\}$.
  Let $\Done \coloneqq \{\, \Gamma \mapsto m, \Delta \mapsto n \,\}$,
  $\Dtwo \coloneqq \Done \cup \{\, \Delta^{\top} \mapsto l, \Delta^{\bot} \mapsto (n-l) \,\}$,
  and $\Dthree \coloneqq \Dtwo \cup \{\, \Gamma' \mapsto (m-1) \,\}$ denote
  versions of $\mathsf{D}$ at various points throughout the algorithm's
  execution. Here, $l$, $m$, and $n$ are all arbitrary names generated by
  \newDomainName.

  \Cref{fig:functioncalls} shows the outline of function calls and their return
  values. For simplicity, we refer to FCG nodes by their types. When \visit
  calls \actuallyVisit and returns its output unchanged, the two function calls
  are shortened as a single call to $\visit/\actuallyVisit$.
